---
title: "Presenting a Categorical Variable"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaicData)
library(tidyverse)
library(ggformula)
library(mosaicCore)
knitr::opts_chunk$set(echo = FALSE)
load("Montana_poll.Rda")
Montana_raw <- Montana_poll
Short_answer <- 
  tribble( ~ answer, ~ ans,
           "I am not likely to vote in this election", "unlikely_voter", 
           "Republican Greg Gianforte", "Gianforte",               
           "Democrat Rob Quist", "Quist",  
           "Libertarian Mark Wicks", "Wicks")    
Montana_poll <- 
  Montana_raw %>%
  left_join(Short_answer)
options(digits = 4)
```

## Use data tables!

You know this sort of textbook problem:

> *In a poll of 573 randomly selected people, 210 respondants indicated support for candidate A. What is the 95% confidence interval on the proportion of the population who support candidate A?*

In StatPREP, we take the position that 573 and 210 are not data; they are a *presentation* of data. We happen to have the [actual data](https://surveys.google.com/reporting/survey?survey=mqbskxyva7gb4h4uwgc7j6h2na), which relate to a 2017 Montana special election for its Congressional seat. (We found the data by following the link, which was published [in this story at fivethirtyeight.com](https://fivethirtyeight.com/features/fake-polls-are-a-real-problem/).)

The data are stored in R under the name `Montana_poll`. Take a look at them by running the following command:

```{r show_data, exercise = TRUE}
Montana_poll
```

This lesson introduces you to tools for producing presentations in the form of counts and proportions from data tables. 

## Presentations of counts

Data presentations can be graphical or tabular. The next command block contains a command for making a simple graphic. Note that the command follows the usual goal/formula/data *template*:

- The goal is `gf_counts`.
- The formula is `~ answer` -- a one-sided formula since we are working with only a single variable.
- The `data` argument is `Montana_poll`.

```{r count_graph, exercise = TRUE}
gf_counts( ~ answer, data = Montana_poll)
```

Not so beautiful; the answer categories are so verbose that they overlap. This is often the case with survey data. Systems such as Survey Monkey or Google Forms encourage such a style. 

To simplify the presentations, we've *recoded* the original answer categories into a short form in the variable `ans`. Go back and remake the plot using `ans` instead of `answer`.

## Numerical presentations

If you prefer a presentation using numbers instead of graphical bars, you merely need change the *goal* of the command to `counts()`.

```{r with_counts, exercise = TRUE}
..goal..( ~ ans, data = Montana_poll)
```

Rather have the presentation as proportions? Give another argument to convey this detail: `format = "proportion"`. Rather have it as percent? Use `format = "percent"`.

## A hint of things to come 

When there are multiple variables available, it's natural to wonder how one thing depends on another. For instance, suppose you want to know if men and women differ in their preferred candidates. This involves changing the *formula* argument in the template: the goal and data remain the same.

```{r by_sex, exercise = TRUE}
..goal..( ~ ans | gender, data = Montana_poll)
```

You can use `gf_counts()` or `counts()` with this new formula. For `counts()`, you can also use the `format =` argument to specify proportions or percent.

Since this lesson is about counts of different categories in a *single* variable, we won't dwell on the structure of the formula beyond saying that the `|` means "given" or "conditioned on." That is, `~ ans | sex` means to find the counts of `ans` separately for the different sexes. 

## IN DRAFT: Other lessons

For another lesson document.


### Recoding

Interested in how to recode a categorical variable? The key is to make a new recoding data table with one line for each level of the original code. Like this:

```{r recode, exercise = TRUE}
Short_answer
```

Note that the variable name for the original code is the same as in the original data. The new code has a new name.

You can probably imagine the tedious work of recoding: using `Short_answer` as a reference, go down every row of the original polling data and add in the new code. 

This is an exercise in *data wrangling*. Better to have the computer do it. The complete statement is:

```{r}
Montana_poll <-
  Montana_raw %>%
  left_join(Short_answer)
```

### Inference

Do the men and women differ in their support for Gianforte

```{r}
summary(lm(ans == "Gianforte" ~ gender, data = Montana_poll))
```
