---
title: "Using R in StatPREP"
output: learnr::tutorial
tutorial:
  id: "StatPREP/Intro_to_learnr"
  version: 0.1
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(checkr)
library(ggformula)
library(mosaicCore)
library(mosaic)
library(openintro)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = checkr::check_for_learnr)
```

## StatPREP Lessons

Most StatPREP lessons involve computing with data. Usually, when computing is involved we think about a stand-alone computing package such as Minitab or R/RStudio that is run on it's own. But in StatPREP lessons, the computing platform is fully integrated into the lesson.  

This document shows the kinds of ways lessons can interact with computing. In other words, this document is a lesson showing how lessons work. 

## Your students and computing

Instructors are understandably concerned about using computing with their students. 

* How will my students access the necessary software?

    StatPREP lessons are available to you and your students through an ordinary web browser. They can be accessed on a laptop, a tablet, or even a smartphone.

* How much do I have to teach my students about computing before they can start the lessons.

    Many of the lessons can be followed with no previous background in computing. Usually, such lessons are designed to help students learn about computing while they are undertaking a statistical task. Some lessons focus on the computing itself, e.g. making graphics, accessing data. 

The next sections of this lesson give examples of three styles of computing activities:

1. Interactive app-like activities that don't require any typing of commands.
2. Just-run-it activities that don't require any typing of commands, but where you are welcome to experiment by modifying commands.

Both (1) and (2) can be used by students with no computing experience.

3. Construct a command. These are the most demanding, since students will need to construct proper commands. Some StatPREP lessons including these activities provide scaffolding to help students get the work done and some even give feedback to help guide students to a correct command.

## 1. Interactive apps

As an example of a no-typing computation, this little app plots out the distribution of any of several variables that the US Census uses to describe each of the counties in the US. 


```{r child="Distribution_app.Rmd", context = "server"}
```

----------------------

Note that this app displays the command that generates the graphics. Such displays are meant to help students start to see the structure of commands and what aspects of the commands correspond to features of the output.


## 2. Just run it!

The second kind of activity involves an R-language command that has been pre-formed. You just need to press the "Run code" button to get the result. This command displays the documentation for the `countyComplete` data used in the previous section.

```{r county-help, exercise = TRUE}
help(countyComplete)
```

Although running such commands requires little effort by the student, we think it is still worthwhile to have the student run the command rather than just presenting the result. First, this is an easy way to engage the student. Second, the student has the possibility of modifying the command if she wants to explore. For instance, change the above command to show the documentation for the `Galton` dataset, which was also available in the previous section's app.

Another example: a scatter plot showing the relationship, on a county-by-county basis, of the fraction of the population that is under age 5 and the fraction that is female.

```{r county-scatter, exercise = TRUE, warning = FALSE, message = FALSE}
gf_point(age_under_5 ~ female, data = countyComplete, alpha = 0.1) 
```

Such a graph can be the basis of a class discussion. For instance, what do you think is going on with the counties that have a population that is $< 45$% female? Or ... What might be different about the counties with a large fraction of the population younger than 5? What would you look at to check your hypothesis?

Feel free to explore relationships by changing the command to show any pair of variables of interest to you. Can you find a pair that has a strong negative correlation? (Hint: Read through the documentation and look for variables that your intuition tells you should be negatively correlated.) 

## 3. Constructing commands

As students become more comfortable with computing, you may decide to use lessons that exercise their computing skills directly by having them construct all or part of a command. The command blocks for such activities will often have blanks to fill in or items to change. They will also have a "Submit" button that can be used to generate feedback.

For instance ... The following command will generate a scatterplot when you press "Run code." Your task is to modify the command so that the plot also shows a smooth linear fit to the data. What you need to know:

* The scatter plot is one *layer* of the graphics. You are going to use another layer with `gf_smooth()` which also generates a layer.
* Connect the layers with the `%>%` symbol, called a "pipe."
* When there are two layer-generating functions joined by a pipe, the second inherits the information you gave as arguments to the first.

```{r two-layers, exercise = TRUE}
gf_point(age_under_5 ~ female, data = countyComplete) %>% ..second_layer..
```

```{r two-layers-check, warning = FALSE, message = FALSE}
#USER_CODE <- "gf_point(age_under_5 ~ female, data = countyComplete) %>% gf_smooth()"
#USER_CODE <- "gf_point(age_under_5 ~ female, data = countyComplete) %>% gf_point(color = \"red\")"

code <- for_checkr(USER_CODE)
code <- line_chaining(code, message = "Remember that a command producing two layers will look like `layer1(...) %>% layer2(...).")
check_blanks(code, quote(gf_point(...) %>% ..two..),
                    passif(two == quote(gf_smooth()), "OK, you put in the smoother. But notice that it's not a straight-line fit, as asked. To tell `gf_smooth()` to fit a straight line, give it the argument `method = \"lm\"`. Try again!"),
                    passif(two == quote(gf_smooth(method = "lm")), "That's right."),
                    failif(TRUE, "You want to pipe into `gf_smooth(), not `{{two}}`.")
)
```

