---
title: "Basic Descriptive Statistics"
output: 
  learnr::tutorial:
    progressive: false
tutorial:
  id: StatPREP/BasicDescriptiveStats
  version: 0.1
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(mosaic)
library(ggformula)
library(mosaicData)
library(mosaicCore)
library(NHANES)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = checkr::check_for_learnr)
```

## Organizing statistics

The computer makes it easy to calculate any statistic involved in an introductory statistics course. Some of the most widely used functions in R have to do with basic statistics: `mean()`, `sd()`, `median()`, and so on. It may seem strange at first, but we're going to discourage you from using these functions. Why? Because generally data analysis doesn't rely on a single statistic, but statistics in a context provided by other statistics. Some examples: interpreting the mean is often done in the context of the standard deviation; comparing the mean and median is a good way to look for skew distributions.

Instead of single-purpose functions such as `mean()` and `sd()`, we're going to use a more general-purpose function, `df_stats()`, that organizes the results systematically, so you can calculate all the statistics of interest in one go. Among other things, `df_stats()` works with data frames and makes it easy to break down statistics by group.

Here's an example, finding the mean height of the children recorded in Galton's data.

```{r df_stats0, exercise = TRUE}
df_stats( ~ height, data = Galton, mean)
```

This command uses the standard template for statistics, graphics, and modeling commands: a goal (`df_stats()`), a formula identifying the variable of interest (`~ height`), a data argument specifying the source of the data (`Galton`), and additional arguments giving additional specifications for the calculation.

Possibly you are thinking that the goal should be "calculate the mean," but remember that statistics are generally used in the context of other statistics.  For instance:

```{r df_stats1, exercise = TRUE}
df_stats( ~ height, data = Galton, mean, sd)
```

The `df_stats()` function is set up so that it will give you a statistical summary of a variable by default. To invoke this default behavior, end the statement after the data argument, without listing any statistics to calculate. (Try it in the above chunk.)

You can use the formula in `df_stats()` to specify one or more variables to use in breaking down the statistics. For instance:

```{r df_stats2, exercise = TRUE}
df_stats( height ~ sex, data = Galton, mean, sd)
```



### Reminder about formulas

Recall that there are two main types of formulas:

* One-sided formulas, like `~ age`
* Two sided formulas like these:
    - `wage ~ educ`
    - `outcome ~ age + smoker`
    
There are **no** formulas that look like `wage ~` or just `wage`. So if there is just one variable under consideration, you will always put it on the right-hand side of the formula.

### Exercise

The `KidsFeet` data table contains measurements of foot length and width for 3rd and 4th graders.

Look at the codebook to see what variables are included.

```{r kidsfeet0, exercise = TRUE}
help(KidsFeet)
```

Use `df_stats()` to construct a data table which will break down the mean width into four groups, girls whose `biggerfoot` is `"R"`, boys whose `biggerfoot` is `"R"`, girls whose `biggerfoot` is `"L"`, and boys whose `biggerfoot` is `"L"`.

```{r girls-bigger, exercise = TRUE, exercise.cap = "Create a statistical summary of foot width"}
df_stats(width ~ ____ + ____, data = KidsFeet, mean)
```

<div id="girls-bigger-hint">For the right-hand side of the formula, figure out which variables are needed to specify girls with a right biggerfoot.</div>

## What stats are available?

The `df_stats()` function is a clearinghouse for calculating many different numerical summaries of a quantitative variable. This consolidation helps to guarantee a consistent form of input and output. Many different descriptive statistics are available.

The default set. These will be calculated if you don't give any specific statistic as an argument:

- `mean`
- `sd` -- the standard deviation
- `length` -- the number of cases
- `median`
- `Q1` -- the boundary of the first quartile
- `Q3` -- the boundary of the third quartile
- `min`
- `max`
- `missing` -- how many of the `n` cases have a missing value

You can also compute
- `coverage` -- giving the range of the central 95% of the variable.

And these inferential statistics are available:

- `ci.mean` -- Confidence interval on the mean
- `ci.median` -- Confidence interval on the median
- `ci.sd` -- Confidence interval on the standard deviation
- `coverage` -- The range covered by the central 95% of the data.

If you wish to use a confidence level or coverage level different from 95%, specify the level as an argument, e.g., `ci.mean(level = 0.80)`.

You can also use `df_stats()` to calculate proportions. The functions `prop()` and `ci.prop()` can be used. They take an argument `success = ` specifying what you want the proportion of. For instance, to calculate the proportion of females in different sectors of the economy, `df_stats(~ sex, data = CPS85, prop(success = "F"))`


```{r tidy_stats_qstats_default}
quiz(
  question("Which of the following statistics is *not* in the default set produced by `df_stats()`?",
    answer("the median"),
    answer("the number of cases"),
    answer("the interquartile interval", correct = TRUE),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question("Which of the following is *not* a statistic provided by `df_stats()`?",
    answer("ci", correct = TRUE),
    answer("coverage"),
    answer("ci.median"),
    answer("ci.sd"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

## Designing a lesson

The `NHANES` data frame contains nutrition and health variables for 10,000 people. Let's imagine that you are teaching a class with many nursing students. Design a lesson for these students in which you ask them to compare outcomes of medically relevant variables (say, body mass index (`BMI`) or systolic blood pressure (`BPSysAve`)) as they differ among various groups as defined by, e.g., gender (`Gender`), age (`AgeDecade`), or lifestyle variables (e.g. `SleepTrouble`, `SmokeNow`.) The point of the lesson should be to have your students discover that even if the means or medians clearly differ between groups, this does not mean that every individual can be expected to follow the same pattern. 

Use the command box below to try out commands involving `df_stats()` that will help make the point of your lesson.

```{r lesson0, exercise = TRUE, exercise.cap = "Commands for your lesson"}


```


If you're not satisfied to use the variables mentioned above, you can see the entire list with the command `help(NHANES)`.

## Graphing the output

It's common when presenting grouped statistics to display them graphically. The output of `df_stats()` is arranged to make this relatively easy to do. To illustrate, here is the calculation of the median wage of workers recorded in a 1985 Current Population Survey study, excerpted in the `CPS85` data table. We'll break down the calculation by the sex of the worker and the sector of the economy in which he or she worked.

```{r cps1, exercise = TRUE}
df_stats(wage ~ sector + sex, data = CPS85, median, ci.median)
```

Suppose you want a graphic comparing the median wages of the two sexes for each of the employment sectors. Since the output of the `df_stat()` calculation is a data frame, we can use all the standard graphics tools for graphing that output.

```{r cps2, exercise = TRUE}
df_stats(wage ~ sector + sex, data = CPS85, ci.median) %>%
gf_errorbar(ci.median_wage_lower + ci.median_wage_upper ~ sector,
            color = ~ sex, 
            position = "dodge", width = 0.3) %>%
  gf_labs(y = "Median wage (95% CI)")
```

