---
title: "Histograms and better"
resource_files:
- dependencies.R
runtime: shiny_prerendered
output: learnr::tutorial
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(mosaicData)
library(ggformula)
knitr::opts_chunk$set(echo = FALSE)
```

## Many displays of distribution

"How do I make a histogram?" This is one of the most common early questions we get from statistics instructors starting to work with data. We think this is a wrong way to frame things; the point is not to make a histogram but to display the distribution of a quantitative variable. And, very often, the point is to *compare* the distribution of a quantitative variable across two or more groups. There are many formats for displaying distribution that are effective for comparing two or more groups. A histogram isn't the best.

So, even though this lesson is about making displays that involve only a single quantitative variable, we're going to put that in the context of what you'll eventually have to do when you get to comparing distributions.

## The variable along the horizontal axis

Many displays of distribution represent the variable as position along the *horizontal* axis. The vertical axis is used to represent some calculated quantity, not a variable.

### A single variable

As always with calculations involving a single variable the formula component the command template is

<div style="font-size: 3em;">~ x</div>

For our example, we'll using the data collected by Francis Galton as part of his pathbreaking 19th century examination of the heritability of height. You can display the data with the simple command `Galton`. Or, if you want the documentation, give the command `help(Galton)`.

Each row in the data is one adult. 

```{r galton-data, exercise = TRUE}
Galton
```



The following command block contains an example of a distribution display with the variable on the x-axis.

```{r x-axis, exercise = TRUE}
gf_density( ~ height, data = Galton, fill = "lightgreen")
```

Perhaps you'd prefer another graphics modality, maybe one that doesn't require you to explain what the quantity on the vertical axis means.  Here are some possibilities:

* `gf_ash()` whose name refers to "average shifted histograms"
* `gf_histogram()` where the vertical axis is in counts.
* `gf_dhistogram()` a histogram, but with the vertical axis as a density. 

You can dress up the plots a little with these optional arguments:

- `color = ` where you give the name of a color, e.g. `"lightgreen"`
- `fill = ` which is much like `color` but controls the fill color
- `alpha = ` a number between 0 and 1 setting the transparency. 

### Two variables

Just so that you can see where things are heading, look at the graphics modalities when you are comparing the distribution of height for two different groups: the males and the females.

Since the variable is being plotted on the x-axis, the basic formula is still 

<div style="font-size: 3em;">~ height</div>

There are two basic ways to add `sex` into the plot.

1. Make the formula "height given sex," that is
    
    <div style="font-size: 3em;">~ height | sex</div>

2. Use `color` or `fill` to represent `sex`. This is called "mapping sex to color (or fill)." Since `sex` is a variable, the argument will look like this:
    
    <div style="font-size: 3em;">fill = ~ sex</div>
    
    It's advisable to set `alpha` to a mid-range value, so that you can see both distributions when one is overlaid by another.

```{r height-sex, exercise = TRUE}
gf_density( ~ height, data = Galton, fill = ~ sex, alpha = 0.5)
gf_density( ~ height | sex, data = Galton)
```

Of course, you can use `gf_ash()`, `gf_histogram()`, or `gf_dhistogram()` instead of density. 

## The variable along the vertical axis

In statistical graphics such as scatterplots, the x- and y-axes are used for different variables, with the response variable on the y-axis. In the Galton example, a scatter plot would have height on the y-axis and sex (or mother's height, or whatever). If you introduce distributions with a modality that puts the height variable on the x-axis, students have to reconcile the two different formats, which is perhaps an unnecessary cognitive load.

Here are some examples of commands to create distribution displays with the quantitative variable on the vertical axis:

```{r echo = TRUE, results = "hide", eval = FALSE}
gf_jitter(height ~ sex, data = Galton, width = 0.2, alpha = 0.5)
gf_violin(height ~ sex, data = Galton, fill = ~ sex, alpha = 0.5)
gf_boxplot(height ~ sex, data = Galton, fill = ~ sex, alpha = 0.5)
```

Try them out in the command box:

```{r xydistributions, exercise = TRUE}

```

### Teaching about the displays

For the purpose of helping students understand graphics, it can be useful to show more than one mode at a time, layering them on top of one another. To do this, use `%>%` to "pipe" one graphics layer into another.

```{r ontopof, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton, 
          color = ~ sex, width = 0.2, alpha = 0.5) %>%
  gf_violin(fill = ~ sex, alpha = 0.2)
```

This is quite a long command, so perhaps not one you want to make a learning objective for your students. 

It's worth pointing out that the formula and data components of the first graphics command get inherited by any subsequent commands.

## So many options!

We're offering you a lot of options for displaying distributions. The intention is not that your students should master all of them, but that you should be able to choose.

For instance, you might want to start with histograms, just because those might be the most familiar to you. ("One thing at a time, please!") But as you become more familiar with data-driven statistics, you may start to realize that the two-variable formats are more natural, tell you everything you might get out of a histogram, and address a more central question in statistics: How do groups differ from one another?

