---
title: "Data-science graphics"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggformula)
library(mosaic)
library(mosaicCore)
library(mosaicData)
library(openintro)
library(NHANES)
NHANESsmall <- NHANES %>% 
  select(Gender, Height, Age, Weight, HomeOwn, Depressed) %>% 
  na.omit() %>%
  sample_n(size = 1000)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


## Data-science graphics?

### Background
It might be natural for people who have taught statistics for many years to think that the graphics taught conventionally are the basis for a data-science oriented approach to graphics. I think we can do better by adopting a few conventions that simplify, streamline, and unify graphics. You'll see those conventions in this tutorial and you can read about them in [these notes on stats graphics](https://dtkaplan.github.io/SDS/data-statistical-graphics.html).



## **The graphics frame**

A fundamental decision in designing stat graphics is the choice of variables to define the space in which marks will be placed. That's a rather abstract description, so let's make it much more concrete. 

**You specify the graphics frame by choosing two variables from a data frame**.

- Graphics will *always*¹ have a variable mapped to the y axis and a variable mapped to the x axis.
- Put the response variable on the y-axis.
- If there is just one variable involved, you should rethink whether the graphic is appropriate. If so, then think about there being a second, "explanatory" variable which is a categorical constant, e.g. "all."

You won't have much occasion to make an empty graphics frame, but we'll start with that. We'll use the `Galton` data as an example:

```{r empty1, exercise = TRUE}
gf_blank(height ~ mother, data = Galton)
```

### Exercise: A categorical axis

Make an blank frame to show data on height versus sex from the `Galton` data.

```{r empty2, exercise = TRUE}
gf_blank( ...your.formula.here... , data = Galton)
```

### Exercise: Two categorical axes

The `NHANESsmall` data² frame has many categorical variables, e.g. `Gender`, `HomeOwn`, `Depressed`. Make a blank frame that can be used to display two of these variables. 

```{r empty3, exercise = TRUE}
gf_blank( ...your.formula.here... , data = NHANESsmall)
```

```{r empty3-solution}
gf_blank(HomeOwn ~ Depressed, data = NHANESsmall)
```

### Exercise: Just one variable

You're going to put the one variable on the vertical axis. Since there's no variable for the x axis, we can create one in the formula by putting a suitable character string on the right-hand side of the formula, e.g. `HomeOwn ~ "All together"


```{r empty4, exercise = TRUE}
gf_blank(HomeOwn ~ "All together", data = NHANESsmall)
```

### Exercise: A frame with three variables

Make the frame in the standard way, then pipe the output to `gf_facet_wrap()`, which takes a one-sided formula as an argument.

```{r empty5, exercise = TRUE}
gf_blank(HomeOwn ~ Depressed, data = NHANESsmall) %>%
  gf_facet_wrap( ~ Gender)
```


## **Displaying data**

To plot your data in the frame, use `gf_point()` instead of `gf_blank()`.

```{r points1, exercise = TRUE}
gf_point(height ~ mother, data = Galton)
```

To employ jittering, use `gf_jitter()` instead of `gf_point()`. It's particularly appropriate to jittering when a categorical variable is involved.

```{r points2, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton)
```

### Exercise: Transparency and changing the jitter width

Use the `alpha =` argument to either `gf_point()` or `gf_jitter()` to set the transparency of the points. A value of 1 means "opaque." A value of 0 means "completely transparent" (which isn't very useful in practice.) In between is in between, e.g. 0.2 is pretty transparent, although for very large data sets you might want something like `alpha = 0.01`.

You may not like how broad the jittering zones are. You can control how much the points are jittered with the `height =` and `width =` arguments to `gf_jitter()`. Both are in the units displayed on the graphics frame. For categorical axes, the distance between two adjacent levels is 1.

Add arguments to the next command to make the horizontal jitter width 0.2 and the vertical jitter 0. Add transparency, setting `alpha` to a value that makes pretty clear the relative density of points at different values of height.

```{r points3, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton, ...your.additional.arguments... )
```

```{r points3-solution, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton, width = 0.2, height = 0, alpha = 0.1)
```

### Exercise: Jittering quantitative variables

Sometimes a quantitative variable takes on only a small set of discrete values. This may not be the part of the story that you want to emphasize with your graph. In such situations, it's appropriate to use jittering along a categorical axis. Typically the default values will work well.

Modify the command so that there is a little bit of jittering along the vertical axis. Remember that the units for the `height` and `width` parameters is the same as shown on the axis. 

```{r points4, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton)
```


```{r points4-solution, exercise = TRUE}
gf_jitter(height ~ sex, data = Galton, width = 0.2, alpha = 0.1, height = 0.5)
```

### Exercise: More than two variables?

Make a plot that displays all these variables from `NHANESsmall`: `Gender`, `Age`, `Height`, `HomeOwn`. You will want to pick two variables for the frame. For a categorical variable, you can use the `color = ` argument to `gf_point()`/`gf_jitter()` effectively. Also for a categorical variable, you can pipe to `gf_facet_wrap()` to display that variable in different facets of the graph.



### Comment: Is that all?

Just two commands? Don't you need something other than `gf_point()` and `gf_jitter()`? 

No. Remember, we're trying to streamline and to enforce distinctions among data graphics, graphics of statistics, and graphics for inference.

## **Graphics for descriptive statistics**

Descriptive statistics can be graphed in a distinctive way: lines and curves that are mostly horizontal.

### Means and such

The `df_stats()` function will calculate groupwise statistics. The process for plotting such statistics is

1. Make a data frame with the statistics.
2. Graph out the contents of the data frame using `gf_errorbar()`.

For instance:

```{r means1, exercise = TRUE}
Stats <- df_stats(height ~ sex, data = Galton, mean)
# Look at Stats to make sure you have the right variable names.
gf_errorbar(mean_height + mean_height ~ sex, data = Stats, color = "blue")
```

The formula argument to the `gf_errorbar()` function is of the form 

    bottom_of_bar + top_of_bar ~ explanatory_variable
    
To draw a single horizontal line, we're re-purposing `gf_errorbar()` by setting the bottom and top of the bar at the same level.


#### Exercise: means in context

The previous graphic shows the different height means for males and females, but the axis limits do not put them in any context.  Here are two good ways to do so:

1. Set the vertical axis to show zero, which makes the picture more like looking at people from head to toe. To do this, pipe the output of `gf_errorbar()` to `gf_lims(y = c(0, NA))`. (The point of the `NA` is signal that you want R to figure out an appropriate upper bound. Or, you can set it explicitly by specifying a number.)

```{r means2, exercise = TRUE}
Stats <- df_stats(height ~ sex, data = Galton, mean)
# Look at Stats to make sure you have the right variable names.
gf_errorbar(mean_height + mean_height ~ sex, data = Stats, color = "blue") %>%
  gf_lims(y = c(0, NA))
```

2. Include the actual data rows so that the person-to-person variation sets the scale.

```{r means3, exercise = TRUE}
Stats <- df_stats(height ~ sex, data = Galton, mean)
# Look at Stats to make sure you have the right variable names.
gf_jitter(height ~ sex, data = Galton, width = 0.2, alpha = 0.2) %>%
gf_errorbar(mean_height + mean_height ~ sex, data = Stats, color = "blue") 
```

#### Comment: Why `gf_errorbar()`?

Wouldn't it be nicer to have a `gf_` function specifically for point statistics? Yes.

But we want to discourage you from plotting point statistics. See the section on graphics for inferential statistics.

### Regression

When you have two quantitative variables, regression provides a way to describe a model-relationship between them. Use `gf_smooth()`.

```{r regression1, exercise = TRUE}
gf_smooth(height ~ father, data = Galton)
```

Two things are different from the usual plot of a regression line:

* It's not a line, it's a smooth curve. If you want a straight line, use `gf_lm()` instead. But why? Are we trying to teach students about lines or about relationships?
* Confidence bands are being shown by default. You can suppress them with the argument `se = FALSE`. But why would you want to suppress them. For `gf_lm()` they are not shown by default, but you can add them with the argument `interval = "confidence"`. You can also use `interval = "prediction"`.

#### Exercise: Regression on two variables

You can easily throw a categorical covariate into the regression. Use `color =` to specify the variable, as with `color = ~ sex`. Try it.

```{r regression2, exercise = TRUE}
gf_smooth(height ~ father, data = Galton, color = ~ sex)
```

#### Exercise: Regression with the data

You've probably figured out that you can show the data along with the regression line. Use `gf_point()` to show the data rows, then pipe to `gf_spline()`.  Try it!

```{r regression3, exercise = TRUE}
# Your answer here.
```

```{r regression3-solution}
gf_point(height ~ father, data = Galton, color = ~ sex) %>% 
  gf_smooth(height ~ father, data = Galton, color = ~ sex)
```


### Density

You're thinking about histograms and density plots. Those are conventional graphics, see `gf_histogram()` and `gf_density()` among others. These both take a one-sided formula, because there is only one variable involved. That violates the principle that graphics frames should have two variables.³ And there's a better way: using `gf_violin()`. 

```{r}
gf_violin(height ~ sex, data = Galton)
```

If you're laying the density function over data, better to fill the violin with a transluscent color and suppress the outlines, for instance

```{r density1, exercise=TRUE}
gf_jitter(height ~ sex, data = Galton, width = 0.1, alpha = 0.2) %>%
  gf_violin(alpha = 0.5, fill = "blue", color = NA)
```

#### Exercise: Add a box-and-whiskers plot

You can replace the violin with a box-and-whiskers plot, or add one in over the violin. For box-and-whiskers plots, you do not want to set the color to NA. Make it something visible (like `color = "blue"`). And you don't really need fill, so you can set `fill = NA`.

Modify the command above to replace the violin with a box-and-whisker plot.  Then, use both the violin and the box-and-whisker together.

```{r density2, exercise=TRUE}
gf_jitter(height ~ sex, data = Galton, width = 0.1, alpha = 0.2) %>%
  gf_violin(alpha = 0.5, fill = "blue", color = NA)
```


```{r density2-solution, exercise=TRUE}
gf_jitter(height ~ sex, data = Galton, width = 0.1, alpha = 0.2) %>%
  gf_violin(alpha = 0.5, fill = "blue", color = NA) %>%
  gf_boxplot(fill = NA)
```

#### Comment: Why box and whisker?

Box and whisker plots were invented for a time when computation was expensive and graphics were drawn with a pen. We have new tools: non-parametric density estimation, fill, transparency, jittering. So consider when any real purpose is served by teaching about box-and-whisker plots. 


## **Graphics for inferential statistics**

In the conventional statistics course, statistical inference is often turned into "find the p-value." But the American Statistical Association has called for a [de-emphasis of p-values](https://www.amstat.org/asa/News/ASA-P-Value-Statement-Viewed-150000-Times.aspx).

### Means and such

Use `df_stats()` to calculate the confidence intervals and `gf_errorbar()` to display them.

```{r infmeans1, exercise = TRUE}
Stats <- df_stats(height ~ sex, data = Galton, ci.mean)
# Look at Stats to make sure you have the right variable names.
gf_jitter(height ~ sex, data = Galton, width = 0.2, alpha = 0.2) %>%
  gf_errorbar(lower + upper ~ sex, data = Stats, color = "blue")
```

### Regression

Use `gf_smooth()` or `gf_lm()`, making sure to turn the confidence band on. This is the default for `gf_smooth()` and can be accomplished in `gf_lm()` with an argument `interval = "confidence"`. All this was covered earlier.

I'm going to make one small change. `gf_smooth()` and `gf_lm()`, by default, draw the regression curve in blue, emphasizing it. But the point of inferential statistics is to indicate the range of plausible values for the statistics being calculated. So why emphasize just one plausible value? Show the range.  
```{r infsmooth1, exercise = TRUE}
gf_point(height ~ father, data = Galton, alpha = 0.25) %>%
  gf_lm(interval = "confidence", color = "NA")
```
Of course, you can also give a little emphasis to the center of the estimate, e.g. `color = "gray"`.


### Density 

We don't do very much with inference on histograms or densities. Usually, we don't have a definite null hypothesis. The questions of interest to students involve the details of the density distribution: are there two peaks, etc.

Here's a way to talk about inference on densities: Use resampling to generate a host of plausible estimates of the density, and use transparency to emphasize the places where they overlap.

```{r infdensity1, exercise = TRUE}
gf_violin(height ~ sex, data = Galton, fill = NA) %>%
  gf_violin(data = resample(Galton), fill = "blue", color = NA, alpha = 0.10) %>%
  gf_violin(data = resample(Galton), fill = "blue", color = NA, alpha = 0.10) %>%
  gf_violin(data = resample(Galton), fill = "blue", color = NA, alpha = 0.10) %>%
  gf_violin(data = resample(Galton), fill = "blue", color = NA, alpha = 0.10) %>%
  gf_violin(data = resample(Galton), fill = "blue", color = NA, alpha = 0.10)
```

Hard to justify from the available data the little dip at 64 inches for females.

## **Proportions**

We're so used to treating proportions as data that we forget they are actually statistics: numbers that summarize a data frame. As such, use `df_stats()` to calculate them and plot them with `gf_errorbar()`.  We'll do data, statistical, and inferential graphics in one go. For the statistics and inference, we'll take advantage of a trick: the proportion at a given level is the mean of a 0/1 variable.

```{r props1, exercise = TRUE}
With_indicator <- NHANESsmall %>%
  mutate(probability = as.numeric(Gender == "female")) 
Stats <-
  With_indicator %>%
  df_stats(probability ~ HomeOwn, mean, ci.mean)
  gf_jitter(probability ~ HomeOwn, data = With_indicator, 
            alpha = 0.1, height = 0.2, color = ~ Gender) %>%
  gf_errorbar(lower + upper ~ HomeOwn, data = Stats) 
```

This is a complex command by any standard. The result is ugly because the y-axis has sex indicated in a non-standard way. On the other hand, the method generalizes well to other ideas about proportion, like this regression of homeownership on age.

```{r props2, exercise = TRUE}
With_indicator <- NHANESsmall %>%
  mutate(probability = as.numeric(HomeOwn == "Own")) 
gf_jitter(probability ~ Age, data = With_indicator, 
            alpha = 0.1, height = 0.2, color = ~ HomeOwn) %>%
  gf_smooth(color = NA, alpha = 1) %>%
  gf_labs(y = "Probability of owning one's home.")
```

You can hardly be faulted for doing the stats calculation within a conventional graphic. 

```{r props3, exercise = TRUE}
gf_props( ~ HomeOwn, data = NHANESsmall, fill = ~ Gender, position = "fill")
```

### Exercise: Inference on bar charts

We can always do inference with resampling and transparency:

```{r props4, exercise = TRUE}
gf_props( ~ HomeOwn, data = NHANESsmall, fill = ~ Gender, position = "fill", alpha = 0.2) %>%
  gf_props( ~ HomeOwn, data = resample(NHANESsmall), fill = ~ Gender, position = "fill", alpha = 0.2) %>%
  gf_props( ~ HomeOwn, data = resample(NHANESsmall), fill = ~ Gender, position = "fill", alpha = 0.2) %>%
  gf_props( ~ HomeOwn, data = resample(NHANESsmall), fill = ~ Gender, position = "fill", alpha = 0.2) %>%
  gf_props( ~ HomeOwn, data = resample(NHANESsmall), fill = ~ Gender, position = "fill", alpha = 0.2) %>%
  gf_props( ~ HomeOwn, data = resample(NHANESsmall), fill = ~ Gender, position = "fill", alpha = 0.2) 
```


## Footnotes

1. Well, *almost always*. One doesn't want to be too doctrinaire. But the departures from doctrine should reflect necessity and efficiency rather than a simple failure to be consistent.]
2. A simplification of the full `NHANES` with just 1000 rows and no missing data.
3. When we treat "frequency" as a variable, we can easily confuse students.

-->